\section{The Multi-group Neutron Transport Equation}

Section \ref{sec:CE_NTE} presented the derivation and equivalent forms of the \gls{ce} \gls{nte}. If using a deterministic tool to solve the \gls{nte}, using \gls{ce} cross sections over typical energy ranges of \(10^{-3}-10^7\) eV is impractical. Therefore, some type of discretization is required for the energy dependence. While the dependence of the angular flux on angle is typically rather weak such that only a few terms in a function expansion could be used for representation, the dependence of angular flux on energy is a very complicated function of the energy range; at high energies, the energy dependence is dominated by the fission spectrum, while at intermediate energies by resonance absorption and at low energies by thermalization effects. A single set of functions to describe all of this behavior would require very many terms, so a discrete ordinates in energy approach is almost universally used. Deterministic transport solvers use a \gls{mg}-in-energy form of the \gls{nte} that is obtained by averaging the \gls{nte} over all energies. For \(G\) energy groups, this produces \(G\) coupled equations, each a function of space, time, and angle, for \(G\) fluxes. Energy group \(g\) is defined over \(E_{g-1}\) to \(E_{g}\), with \(1\leq g\leq G\). Groups are ordered in order of decreasing energy, equivalent to increasing lethargy. The greatest amount of simplification occurs when using a single energy group, which is equivalent to assuming that energy does not change in scattering collisions.

Several definitions are required for the derivation of the \gls{mg} \gls{nte}. The angular flux of group \(g\) is defined as the angular flux integrated over the energy bounds for group \(g\):

\beq
\label{eq:GroupwiseQuantity}
\psi_g(\vv{r},\hO,t)\equiv\int_{E_{g-1}}^{E_g}dE\psi\seat
\eeq

Groupwise cross sections for reaction \(i\) are defined in a similar manner based on the flux-weighted cross sections:

\beqa
\label{eq:GroupwiseQuantityProduct}
\Sigma_{i,g}(\vv{r},\hO,t)\equiv&\frac{\int_{E_{g-1}}^{E_g}dE\Sigma_{i}\seat\psi\seat}{\int_{E_{g-1}}^{E_g}dE\psi\seat}\\
=&\frac{\int_{E_{g-1}}^{E_g}dE\Sigma_{i}\seat\psi\seat}{\psi_g\sat}\\
\eeqa

where Eq. \eqref{eq:GroupwiseQuantity} has been inserted for the denominator. Using Eqs. \eqref{eq:GroupwiseQuantity} and \eqref{eq:GroupwiseQuantityProduct}, the \gls{mg} \gls{nte} can be derived for group \(g\) by simply integrating the \gls{nte} in Eq. \eqref{eq:nte1} over \(E_{g-1}\leq E\leq E_g\):

\beqa
\label{eq:mg_nte}
&\int_{E_{g-1}}^{E_g}dE\frac{\partial}{\partial t}\left(\frac{\psi\seat}{v(E)}\right)+\hO\cdot\nabla\psi_g\sat+\Sigma_{t,g} \psi\sat=\\
&\hspace{1cm}\sum_{g'=1}^G\dhOprime\Sigma_s^{g'\rightarrow g}(\vv{r},\hO'\rightarrow\hO,t)\psi_{g'}(\vv{r},\hO',t)\ +\\
&\hspace{2cm}\int_{E_{g-1}}^{E_g}dE\promptfissionsource\psi\seatprime +\\
&\hspace{3cm}\sum_{j=1}^J\chi_{d,j,g}\lambda_jC_j(\vv{r},t)+ S_g\sat
\eeqa

where \(\hO\cdot\nabla\) was brought outside the streaming term since this operators is not a function of energy.

The following sections describe the generation of group constants for energies in the fast and thermal ranges. These methods are based on approximations that neglect spatial dependence and use theory to predict the detailed energy spectrum over which the microscopic cross section data can be averaged.

\subsection{Thermal Spectrum Group Constants}
This section describes the theory behind the generation of group parameters for energies in the thermal range. The generation of thermal group constants is complicated by the complex neutron scattering process at low energies, where the thermal motion of the nuclei cannot be neglected, and neutron energies are typically on the order of binding energies of the nuclei such that interaction tends to occur between a neutron and a group of nuclei simultaneously. Scattering cross sections are strongly dependent on both energy and angle. Further, unlike fast energy cross sections that tend to only depend on the identity of the interacting media, thermal cross sections depend strongly on the state of matter and temperature of the interacting medium.